---
title: "Data types"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
In this task we use some basic R commands for working with a dataset. These include importing, inspecting the size, inspecting the data types and inspecting the variable types. Additionally a function is created that summarises categorical and numerical variables in a dataframe.

# Task steps
## Define a string
This section defines a string containing my name, the unit name and the task name.
```{r}
description = 'Darrin William Stephens, SIT741: Statistical Data Analysis, Task T01.P1'
```

Print the string using the cat() function.
```{r}
cat(description)
```

## The dataset
The dataset I chose for this task was the blood pressure measurement data from the US Centers for Disease Control and Prevention						
National Health and Nutrition Examination Survey. It contains three categorical and nine numerical variables.

Based on 2017-March 2020 Blood Pressure - Oscillometric Measurement (P_BPXO).
Published: May 2021						
Description: https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/P_BPXO.htm						
License: public domain as per https://www.cdc.gov/other/agencymaterials.html

Information about dataset variables is shown in the table below.

|Variable name | Data type| Description |
|:-------------|:---------|:------------|
|SEQN	         |Numeric	  |Respondent sequence number|
|BPAOARM	     |Character	|Arm selected - oscillometric|
|BPAOCSZ	     |Numeric		|Coded cuff size - oscillometric|
|BPXOSY1	     |Numeric		|Systolic - 1st oscillometric reading|
|BPXODI1	     |Numeric		|Diastolic - 1st oscillometric reading|
|BPXOSY2	     |Numeric		|Systolic - 2nd oscillometric reading|
|BPXODI2	     |Numeric		|Diastolic - 2nd oscillometric reading|
|BPXOSY3	     |Numeric		|Systolic - 3rd oscillometric reading|
|BPXODI3	     |Numeric		|Diastolic - 3rd oscillometric reading|
|BPXOPLS1	     |Numeric		|Pulse - 1st oscillometric reading|
|BPXOPLS2	     |Numeric		|Pulse - 2nd oscillometric reading|
|BPXOPLS3	     |Numeric		|Pulse - 3rd oscillometric reading|

I have used this dataset for a previous Deakin unit and had cleaned (missing values removed) CSV file available.

## Import the dataset
The class of the columns is set on import. The first three are specified as categorical (factor) and the next nine are specified as numeric.
```{r}
data = read.csv('Blood_Pressure_Measurement_cleaned.csv', colClasses=c(rep("factor", 3), rep("numeric", 9)))
```

## Inspect the head of the dataset
To inspect the header and the first n columns of the dataset, I pass the dataframe to the head() function. As I don't specify the number of rows to show the default number of rows (6) is shown.
```{r}
head(data)
```

## Inspect the size of the dataset
To get the number of observations and the number of variables I use the dim() function. This returns the dimensions of the dataframe.
```{r}
# Get the dimensions of the dataset
dims = dim(data)
obs = dims[1]
vars = dims[2]

# Print the values
cat("Number of observations:", obs, "\n")
cat("Number of variables:", vars, "\n")
```

## Inspect data types
To inspect the data type I use the sapply function to apply the typeof function to each column in the dataframe. The typeof function returns the data type for each column (variable).
```{r}
sapply(data, typeof)
```


## Function to summarise
A function that takes a dataframe and summarises the data within the dataframe. The function loops through all columns of a dataframe and 
for each column it checks if the column is categorical or numerical. For categorical columns a frequency table is produced and for numerical columns the mean is calculated. The summary infrmation for each column is printed.

```{r}
summarise = function(df){
  # This function loops through all columns of the provided dataframe
  # For each column it checks if the column is categorical or numerical
  # If categorical a frequency table is produced
  # If numerical the mean is calculated
  for (col_name in names(df)){
    # Print column name
    cat("Variable:", col_name)
    
    # Check if categorical 
    if (is.factor(df[[col_name]])){
      cat(" is categorical.\n")
      cat("Frequency table:")
      print(table(df[[col_name]]))
      cat("\n")
    }
    
    # Check if numerical
    if (is.numeric(df[[col_name]])){
      cat(" is numerical.\n")
      cat("Mean:", mean(df[[col_name]]), "\n")
      cat("\n")
    }
  }
}
```

## Summarise the imported dataset
The first column (SEQN) of the imported dataset is a nominal categorical variable that represents an identification number for each person in the survey. Since identification numbers are uniquie a frequency table for this variable would consist of 9410 ones. This is a lot of information to print, therefore, I have removed the first column from the dataframe for this section. The code below shows creating a new dataframe by excluding the first column from the original dataframe. Each column (variable) in the new dataframe is then summarised by using the summarise function.

```{r}
new_data = data[-c(1)]
summarise(new_data)
```
